A Large-scale Study of Polymorphism in R

### 2019-04-02 paper/main.tex
levels

\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt #1}}\xspace}     % Any

@@

The logical ``baseline'' for the design of a type system for R is to adopt
the language's dynamic types (in the sense of \code{typeof}).  We call this
L0.  The types reflects the types of values in the R runtime environment, as
the \code{typeof} an object is in direct correspondence with the runtime
type tag of the value.

Thus L0 includes the following types: vectors of primitive types (integers
\I, double \D, character \C, logical \L, raw \R, complex \X), lists of any
type (\l), and scalar values (environment \sE, closures \sF, symbols \sY, S4
object \sS, and null \sN). We add the type \ANY to stand in for any type.

We also include a notion of subtyping that follows from the conversion rules
of the language.  All types are subtypes of \ANY, thus {\tt T}<:\ANY.
Moreover, logicals are subtypes of integers, which are subtypes of doubles,
\L <: \I <: \D.

@@

L0 is limited in expressivity. To address its limitations, we augment it
scalars, parametrized lists.  In addition, we collect attribute and class
information to distinguish a type for data frames and matrices, as the
\code{typeof} these is simply {\tt list} in R.

The second level enhances {\bf L0} with scalar values for primitive types
(\sI, \sD, \sC, \sL, \sR, and \sX), type parametric lists (written \lT{T}
for any type {\tt T}), data frames \attr{\l}{df}, and matrices
\attr{\vec{\tt T}}{mat}.

\subsection{L2 Types}
 {\bf L2}: The third level adds attributes lists...

While we do consider some class and attribute information when constructing
our types, it is also valuable to consider them entirely separate from
types.  Users are free to define their own classes, and arbitrarily add and
remove attributes, and interesting patterns might arise in considering them.
That said, not attribute patterns and classes are necessarily interesting,
and indeed some of R's built-in data structures use them to some extend.
For one, data frames have a {\tt data.frame} class (explicitly on the class
attribute).  Other standard patterns are:

\begin{itemize}
\item data frames have three attributes: {\tt names} for column names, {\tt
  row.names} for row names, and {\tt class} (which is instantiated to {\tt
  "data.frame"} upon creation).
\item matrices have at least a {\tt dim} attribute with matrix dimensions,
  and optionally a {\tt dimnames} attribute for dimension names. Curiously,
  the {\tt dim} attribute is all that is needed to define a matrix (i.e., a
  list with a {\tt dim} attribute {\it is} a matrix).
\item lists and vectors may have a {\tt names} attribute, which assigns
  names to locations.
\item to ease dealing with large sorted sets of non-numeric values, R offers
  the \code{factor} and \code{levels} functions.  \code{factor(x,
    levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which
  specifies an ordering for \code{x} according to the \code{sortedOrder}
  parameter.  \code{factor} also changes the class of the parameter being
  factored to \texttt{"factor"}.
\end{itemize}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 29 18:10:23 2019
\begin{table}[ht]\label{fig:L1argcounts}\centering\begin{tabular}{lrr}  \hline
Type                   & Count & Percentage \\  \hline
Full Monomorphic       & 784K & 85.76 \\
  Monomorphic in Type & 845K & 92.46 \\
  Monomorphic in Class & 827K & 90.41 \\
  Monomorphic in Attribute Pattern & 825K & 90.27 \\
  Total Seen & 914K & --- \\
   \hline
\end{tabular}
\caption{typeof type system (L0) (arguments)}
\end{table}
% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 29 18:10:23 2019
\begin{table}[ht]
\label{fig:L1funcounts}
\centering
\begin{tabular}{lrr}
  \hline
Type & Count & Percentage \\
  \hline
Full Monomorphic & 144K & 66.54 \\
  Monomorphic in Type & 173K & 80.16 \\
  Monomorphic in Class & 163K & 75.21 \\
  Monomorphic in Attribute Pattern & 161K & 74.37 \\
  Total Seen & 216K & --- \\
   \hline
\end{tabular}
\caption{typeof type system (L0) (functions)}
\end{table}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsL1}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\
  \hline
  Full Monomorphic & 143K & 65.88 \\
  Monomorphic in Type & 167K & 77.25 \\
%  Monomorphic in Class & 163K & 75.24 \\
%  Monomorphic in Attribute Pattern & 161K & 74.37 \\
  Total Seen & 216K & --- \\
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L1.}
\end{table}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsL1p}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Polymorphism & Count & Percentage \\
  \hline
  Full Monomorphic & 781K & 85.47 \\
  Monomorphic in Type & 869K & 95.01 \\
%  Monomorphic in Class & 831K & 90.31 \\
%  Monomorphic in Attribute Pattern & 830K & 90.25 \\
  Total Seen & 914K & --- \\
     \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in L$1^{+}$.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsL1p}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\
  \hline
  Full Monomorphic & 143K & 66.04 \\
  Monomorphic in Type & 185K & 85.30 \\
%  Monomorphic in Class & 163K & 75.13 \\
%  Monomorphic in Attribute Pattern & 161K & 74.31 \\
  Total Seen & 216K & --- \\
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L$1^{+}$.}
\end{table}

@@

In L2, we aim to incorporate R's other reflective capabilities to address some of the more subtle instances of polymorphism that we have observed.

First, we reiterate that there are a finite number of results from the {\tt typeof} function, and this is not the case for \code{class} and \code{attributes}.
Classes and attributes can be arbitrarily defined by the user, and users create their own classes and assign their own attributes frequently.
Unlike types (in the sense of \code{typeof}), the number of possible classes and attributes is arbitrary.

Of course, not all classes are user-defined, and R has a number of built-in or otherwise standard classes, ranging from truly primitive classes such as ``numeric'' to the more complex ``data.frame''.
We have already used these when building our more refined types, and so we transform each of R's standard classes into a ``primitive'' class, which will minimize their impact and thus highlight user-defined classes.

As for attributes, again some patterns arise naturally, for instance all matrices have a ``dims'' attribute (indeed, it's this attribute that makes a matrix a matrix).
\AT{This sounds like it's implying that we should do something about that, but I'm not sure if that's a good idea.}
In principle, attributes are name, value pairs, but to keep things simple we will reduce these to name, type pairs, so the name of an attribute and the type that it was observed to have.
We will show an L0 type for attributes.

>>>>

\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
\newcommand{\attrclass}[3]{\ensuremath{#1^{\mathtt{#3}}_{\mathtt{#2}}}\xspace}
\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt #1}}\xspace}     % Any

@@

The logical ``baseline'' for the design of a type system for R is to adopt
the language's builtin types.  Types in this system, which we call {\bf L0}
reflect the types of values in the runtime environment, the \code{typeof} an
object is in direct correspondence with the runtime type tag of the value.
Thus we include the following types: vectors of primitive types (integers
\I, double \D, character \C, logical \L, raw \R, complex \X), lists of any
type (\l), and scalar values (environment \sE, closures \sF, symbols \sY, S4
object \sS, and null \sN). We add one type stand in for unevaluated values,
the type any (written \ANY). In {\bf L0} subtyping follows from the
conversion rules of the language.  Logicals are subtypes of integers, which
are subtypes of doubles, \L <: \I <: \D.  All types are subtypes of \ANY,
thus {\tt T}<:\ANY for an type {\tt T}.

@@

{\bf L0} is limited in expressivity as it supports polymoprhism for list in
a trivial way, it does not address some of the popular constructed data
types such as matrices and data frames, and cannot differentiate between
scalar values and vectors.  {\bf L1} types build on and extend the previous
level with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR,
and \sX), type parametric lists (written \lT{T} for any type {\tt T}), data
frames (written \attr{\l}{df} as they are built up from lists of type any),
and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
type {\tt T}. There is no subtyping relation between lists of different
types.

\subsection{L2 Types}

{\bf L2} extends the previous system with more support for attributes.
While there is a finite number of results from the {\tt typeof} function,
this is not the case for \code{class} and \code{attributes}.  Classes and
attributes are user-defined.  Of course, not all classes are user-defined,
and R has a number of built-in or otherwise standard classes, ranging from
truly primitive classes such as {\tt numeric} to the more complex {\tt
  data.frame}.  We have already used the latter in {\bf L1}. We will ignore
those classes and thus highlight user-defined ones.  As for attributes,
again some patterns arise naturally, for instance all matrices have a {\tt
  dims} attribute (indeed, it's this attribute that makes a matrix a
matrix).  Attributes are name-value pairs, to keep things simple we reduce
these to names. The only attribute that we treat specially is the {\tt
  class} for those we will recurse and capture the set of values defined by
the user (these are the class names for the object).  Thus, we extend {\bf
  L1} with attribute lists and classes: \attrclass{T}{C}{N} where {\tt T}
can be any L1 type other than data frame or matrix, {\tt C} and {\tt N}
stands for a list names.  The subtype relation is extended with
\attrclass{T}{C}{N}<:\attrclass{T}{C'}{D'} if {\tt C} $\subseteq$ {\tt C'}
and {\tt N}$\subseteq${\tt N'}.

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 29 18:10:23 2019
\begin{table}[ht]\label{fig:L1argcounts}\centering\begin{tabular}{lrr}  \hline
Type                   & Count & Percentage \\  \hline
  Monomorphic in Type & 845K & 92.46 \\
  Total Seen & 914K & --- \\
   \hline
\end{tabular}
\caption{typeof type system (L0) (arguments)}
\end{table}
% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 29 18:10:23 2019
\begin{table}[ht]
\label{fig:L1funcounts}
\centering
\begin{tabular}{lrr}
  \hline
Type & Count & Percentage \\
  \hline
  Monomorphic in Type & 173K & 80.16 \\
  Total Seen & 216K & --- \\
   \hline
\end{tabular}
\caption{typeof type system (L0) (functions)}
\end{table}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsL1}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\
  \hline
  Monomorphic in Type & 167K & 77.25 \\
  Total Seen & 216K & --- \\
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L1.}
\end{table}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsL1p}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Polymorphism & Count & Percentage \\
  \hline
  Monomorphic in Type & 869K & 95.01 \\
%  Monomorphic in Class & 831K & 90.31 \\
%  Monomorphic in Attribute Pattern & 830K & 90.25 \\
  Total Seen & 914K & --- \\
     \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in L$1^{+}$.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsL1p}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\
  \hline
  Monomorphic in Type & 185K & 85.30 \\
%  Monomorphic in Class & 163K & 75.13 \\
%  Monomorphic in Attribute Pattern & 161K & 74.31 \\
  Total Seen & 216K & --- \\
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L$1^{+}$.}
\end{table}

@@
